Imputation of Incomplete Multilevel Data with R

### 2024-02-05 Manuscript/manuscript.tex

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Hanne I. Oberman\\
Methodology and Statistics\\
Padualaan 14\\
Utrecht The Netherlands\\
E-mail: \email{h.i.oberman@uu.nl}\\
URL: \url{https://www.hanneoberman.github.io}\\
\\~
Johanna Muñoz\\
Julius Centre for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~
Valentijn M.T. de Jong\\
Julius Centre for Health Sciences and Primary Care\\
Utrecht The Netherlands\\
\\~
Gerko Vink\\
Julius Centre for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~
Thomas P.A. Debray\\
Julius Centre for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~

@@

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, frame hidden, boxrule=0pt, breakable, enhanced]}{\end{tcolorbox}}\fi

@@

This tutorial is dedicated to readers who are unfamiliar with multiple
imputation. More experienced readers can skip the introduction (case
study 1) and directly head to practical applications of multilevel
imputation under MAR conditions (case study 2) or under MNAR conditions
(case study 3).

@@

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, leftrule=.75mm, toprule=.15mm, colframe=quarto-callout-color-frame, rightrule=.15mm, bottomrule=.15mm, colback=white, breakable, opacityback=0, left=2mm]

@@

Regarding the restrictions imposed by the data, for instance, the use of
cluster indicator variables is restricted in datasets where there are
not many clusters and many observations per cluster (Graham, 2009). The
last restriction is also required when imputations are performed on each
cluster separately. When this restriction cannot be achieved, one can
use an imputation model that simultaneously imputes all clusters using a
hierarchical model (Allison 2002).

Under this hierarchical imputation model, observations within clusters
are correlated and this correlation is modeled by a random effect so the
hierarchical model can be estimated even when there are few observations
per cluster. However, this strategy is best suited for balanced data
(Grund, 2017) and when random effects model is appropriated, i.e.~the
number of clusters is adequate. (Austin,2018).

Here it is important to evaluate the assumptions imposed by the main
model, for instance by using the cluster indicator strategy may lead to
bias estimates when the model is based on a hierarchical model
(Taaljard,2008). Even when an imputation strategy congenial with the
main model is preferred, it is important to consider whether it is
appropriate for the data as a less complex imputation strategies may
also lead to unbiased estimates in certain scenarios(Bailey 2020). For
instance, in causal effect analysis, separately imputation may lead to
smaller bias when the size of the smaller exposure cluster is large,
compared with an imputation model that includes exposure-confounder
interactions. (Zhang,2023).

Below we provide a imputation workflow that can be used in general to
impute cluster data.

\hypertarget{focus-on-the-main-analysis}{%
\subsection{Focus on the Main
Analysis}\label{focus-on-the-main-analysis}}

When dealing with incomplete clustered data, start by looking at your
research questions and planned analysis. Pretend there are no missing
data at first. This will help you understand your research hypotheses,
the main statistical model, and give you insights into your data's
structure (refer to the level table), variable types (e.g., confounders,
auxiliary variables), and other considerations such as variable
relationships, interactions or polynomial terms.

@@

Now, explore what's in your data. Use simple tools like histograms and
QQ plots to understand variable distributions, plausible range of values
and identify errors or outliers. Check interactions between predictors
with scatter plots, look for collinearity with VIF or correlations
plots, this may also help you to identify non-considered relationships
that may affect the main model. You might also consider to test
assumptions related to your response variable such as variance
homogeneity, independence between observations (eg. ACF, variagrams).
This will help you choose an imputation model that suits your data. In
addition, the intraclass correlation coefficient (ICC) can be examined
to assess cluster differences, aiding in the choice between the 2l and
1l methods for imputation.

Next, explore the missingness. Look at the
\textbf{proportion of missing values} in the dataset variables. This
helps find potential predictors and cut down on unnecessary variables in
the imputation model. Doing this can lower the risk of multicollinearity
or computational issues, especially with certain parametric imputation
methods. You can also identify predictors for the imputation model by
using inflow criteria to see connections between missing data in one
variable and observed variables, and outflow criteria to identify
connections between observed values in one variable and missing data in
others.

Check the missing patterns at cluster level, this can help you to select
the most appropiated imputation approach in terms of computational
efficiency (e.g., simpler regression imputation versus FCS in univariate
patterns).

@@

Before diving into imputation, make sure your estimation model can
handle missing data. Sometimes, simpler methods like complete case
analysis might be suffice, especially if your missingness is low
(usually \textless5\%). There are scenarios in which specific Maximum
Likelihood (ML) estimation methods outperform Multiple Imputation (MI)
methods, for instance when the response variable is the sole incomplete
variable, mixed models demonstrate robustness to missing data under the
Missing at Random (MAR) assumption and with a correct
variance-covariance specification.\cite{molenberghs2007}.

@@

Clean up your dataset before the imputation process. Keep initially the
essential variables for your main model, but include extra variables if
needed for specific procedures during analysis (e.g.~confounders on
balancing procedures). Think about adding other useful variables, even
if they're not in the main model, such as instrumental or auxiliary
variables which might boost your parameter estimates, especially if
they're associated to the probability of missingness for some incomplete
variables.

Figure out if you can directly impute incomplete variables by just use
deductive imputation. This involves inferring missing values based on
logicallogical connections between variables. It's especially handy for
variables that depend on each other, like calculating BMI from weight
and height.

Deductive imputation is also useful for getting values for level-1
variables from level-2 ones, like in Individual Participant Data (IPD).
In these situations, you can guess missing information from metadata or
by using or through deduction based on time or protocols. For example,
you could deduce missing test values for patients who have passed away.

@@

Start by choosing the imputation model for each incomplete variable in
your dataset. The mice package suggests methods based on variable types
for non-clustered variables, and for the cluster ones, you can use the
micemd package's find.defaultMethod() function. This function selects
from different 2l imputation methods based on cluster size and the
proportion of missing data in each cluster.

@@

The imputation model can incorporate additional terms compared to the
main model without causing compatibility issues. For instance, it's
recommended to include the outcome variable in the imputation model for
prediction variables \cite{moons2006}. In cases where the outcome is
time-to-event, the Nelson-Aalen estimate of the time to the event should
be added as a covariate in the imputation model {[}REF{]}. Additionally,
including auxiliary variables, even if not part of the main model, can
be linked to the probability of missingness, improving the likelihood of
meeting the Missing at Random (MAR) assumption and enhancing estimation
efficiency \cite{hardt2012}.

@@

\begin{tabular}{ll}
     \toprule
Type & Definition \\
  \midrule
-2  & Cluster variable, in this case the one defined by j index  \\
 1  & Fixed variable, e.g., level-1  $x_{1_{ij}}$ or  level-2 $x_{2_{i.}}$ \\
 2  & Random variable, e.g., $x_{3_{ij}}$\\
 3  & *Fixed variable with cluster mean $x_{4_{ij}}$\\
 4  & *Random variable with cluster mean $x_{5_{ij}}$\\
-3. & Random variable only included on selection model (Heckman model)\\
-4. & Random variable only included on main model (Heckman model)\\
      \bottomrule
\end{tabular}

\begin{itemize}
\tightlist
\item
  3 and 4 type have been advised to used as the inclusion of the means
  of the cluster is beneficial on FCS \cite{mistler2017}
\end{itemize}

Recipes have been proposed for imputing incomplete level-1 and level-2
variables for hierarchical models \cite{buur18} (\(\S\) 7.10), which can
be convenient to follow when dealing with numerous variables
(interaction terms at different levels) and models with many random
effects that are prone to convergency problems due to overspecification
in the imputation model. These rules were designed to ensure
compatibility among the conditionally specified imputation models and
congeniality between the imputation and the main model. However, they
may not be applicable to all 2l imputation methods; for instance, the
2l.2stage imputation methods of the micemd package only allow the
inclusion of random predictor variables (2).

@@

\hypertarget{impact-data}{%
\subsection{IMPACT data}\label{impact-data}}

@@

In this example, we demonstrate a multilevel imputation of random
intercept and random slope model with a continuous response. We utilize
the obesity dataset included in the \texttt{micemd}@ package, a
simulated dataset that emulates an electronic survey in which
individuals are asked to provide information about their weight and
consumption habits in different countries. We simulate data for 5
clusters so that the true values are known. We use the following
variables from the dataset:

@@

We observe that the missing pattern is non-monotonic and quite similar
across the clusters. However, regarding the weight variable, we notice
that is systematically missing in cluster 3. In order to evaluate if we
require a imputation method that accounts for clustering we assess the
Intraclass Correlation

@@

It suggests using `2l.2stage.bin' for the FAV variable and
`2l.2stage.norm' for the age variable. However, after inspecting the age
and weight density plot, we consider modifying its method to
`2l.2stage.pmm'. For the BMI variable, we employ deterministic
imputation.

@@

For these imputation models, it is necessary to specify the prediction
matrix, with the cluster variable labelled as -2 and the predictor
variable measured within clusters labelled as 2, encompassing all
variables. We need to suprime the variable Time as this variable is not
specified in the main model. We also modify the relationship between
BMI, weight and height in the prediction matrix to avoid circular
predictions. Then we proceed to run the imputation model.

@@

@@

After confirming convergence, we proceed to save the results for future
use. We consider the possibility that patients may not have been
selected randomly, which would then have led to a distribution for
weight that does not reflect the weight in the population. It???s likely
that an omitted variable, like self-esteem, could influence this
selection. For instance, individuals with lower self-esteem might have
higher weight values, impacting their willingness to provide honest
information due to embarrassment.

To address this situation, two approaches have been proposed for dealing
with Missing Not at Random (MNAR) data: pattern-mixed models and
selection models. Within pattern-mixed models, methods like the delta
method and more advanced ones like NARFS have been suggested. The
selection model approach includes methods such as the Heckman model,
which can be particularly useful in this case. Several methods,
including those by \cite{galimard2018}, and the recently a Heckman
method designed for two-level data, allow for variations in intercepts
and exposure effects (random intercept and slope) \cite{munoz2023}.

To apply the \textbf{2l.2stage.heckman} method, the weight variable
should be specified as `2l.2stage.heckman' found in the micemd package.
Additionally, the prediction matrix needs modification because this
method involves specifying two equations: one for the outcome,
describing the incomplete variable in terms of partially observed
predictors (in this case, all variables from the main model), and the
other for the selection model, explaining the probability of being
observed based (R) on certain variables. For the outcome equation we
consider the same imputation model that we used for the MAR case (main
model).
\[Weight_{ij}= \beta^O_{o} + \beta^O_{1}Age_{ij} + \beta^O_{2}FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}\]
Regarding the selection equation, we include the same predictors as
those in the main model, as well as a time variable. Here the time
variable serves as a restriction exclusion variable specifically
explaining the probability of being observed but not affecting the
incomplete value (Weight). In this context, we assume that the time a
user spends completing the survey serves as a proxy for the barriers
they may encounter in survey completion, such as familiarity with the
survey content or internet speed. These factors may lead the user to
skip specific questions or even the entire survey. Also, we assume the
time does not have any influence on the subject???s weight.
\[R_{ij}= \beta^S_{o} + \beta^S_{1}Age_{ij} + \beta^S_{2}FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}\]

@@

We note that there is minimal disparity in the age effect, FamObs, or
Gender across the various imputation models under consideration. An
analysis of the intercept reveals that, under the MNAR assumption, a
higher average BMI is anticipated compared to the MAR assumption.
Nonetheless, with respect to precision of estimates, we notice that in
general MNAR imputation leads to wider confidence intervals, in this
case it does not have any influence on the final result but there could
be cases where variation in the assumed missing mechanism could lead
also to differences on significant test and therefore lead to
contradictory conclusions.

@@

The ``Mice'' and ``mice''-based R packages offer a range of imputation
methods tailored for hierarchical data, easily adaptable to the
dataset's structure. Before proceeding with the analysis of the imputed
dataset, it is essential to assess the convergence of the imputation
method. This evaluation can reveal issues such as circular problems, the
need for additional iterations, or challenges associated with the chosen
imputation method.

\hypertarget{funding}{%
\section{Funding}\label{funding}}

@@

>>>>

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Hanne I. Oberman\\
Methodology and Statistics\\
Padualaan 14\\
Utrecht The Netherlands\\
E-mail: \email{h.i.oberman@uu.nl}\\
URL: \url{https://www.hanneoberman.github.io}\\
\\~
Johanna Muñoz\\
Julius Center for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~
Valentijn M.T. de Jong\\
Julius Center for Health Sciences and Primary Care\\
Utrecht The Netherlands\\
\\~
Gerko Vink\\
Julius Center for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~
Thomas P.A. Debray\\
Julius Center for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~

@@

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi

@@

This tutorial is dedicated to readers who are unfamiliar with multiple
imputation. More experienced readers can skip the background section and
introduction (case study 1) and directly head to practical applications
of multilevel imputation under MAR conditions (case study 2,
Section~\ref{sec-impact}) or under MNAR conditions (case study 3).

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, breakable, bottomrule=.15mm, colback=white, left=2mm, colframe=quarto-callout-color-frame, rightrule=.15mm, leftrule=.75mm, arc=.35mm, toprule=.15mm]

@@

Regarding the restrictions imposed by the nature of the data, i.e., the
sample size, the use of cluster indicator variables is restricted in
datasets where there are not many clusters and many observations per
cluster (Graham, 2009). The last restriction is also required when
imputations are performed on each cluster separately. When this
restriction cannot be achieved, one can use an imputation model that
simultaneously imputes all clusters using a hierarchical model (Allison
2002).

Under this hierarchical imputation model, observations within clusters
are correlated and this correlation is modelled by a random effect so
the hierarchical model can be estimated even when there are few
observations per cluster. However, this strategy is best suited for
balanced data (Grund, 2017) and when the number of clusters is adequate
sufficient for the random effects model (Austin,2018).

Here it is important to evaluate the assumptions imposed by the main
model, for instance the cluster indicator strategy bias the estimator
when the model is based on a hierarchical model (Taaljard,2008).
Although an imputation strategy congenial with the main model is
preferred, it is important to consider whether a complex imputation
model is appropriate for the data, as a less complex imputation
strategies may also lead to unbiased estimates in certain scenarios
(Bailey 2020). Below we provide a imputation workflow that can be used
in general to impute cluster data.

@@

First, explore the data, preferably using visual tools. This will help
you choose an imputation model that suits your data. In addition, the
intraclass correlation coefficient (ICC) can be examined to assess
cluster differences, aiding in the choice between methods for dependent
(such as clustering) or independent data.

Next, explore the missingness. Look at the
\textbf{proportion of missing values} in the dataset variables.
Variables for which most data is missing may contribute little in the
analysis, and may be removed from the analysis model. Removing them also
from the imputation model can lower the risk of multicollinearity or
computational issues, especially with certain parametric imputation
methods. You can also identify predictors for the imputation model by
using inflow criteria to see connections between missing data in one
variable and observed variables, and outflow criteria to identify
connections between observed values in one variable and missing data in
others.

Check the missing patterns at cluster level, this can help you to select
the most appropriate imputation approach in terms of computational
efficiency (e.g., simpler regression imputation versus FCS in univariate
patterns).

@@

Before diving into imputation, make sure your main model can handle
missing data. Sometimes, simpler methods like complete case analysis
might be suffice, especially if your missingness is low (usually
\textless5\%). There are scenarios in which specific Maximum Likelihood
(ML) estimation methods outperform Multiple Imputation (MI) methods, for
instance when the response variable is the sole incomplete variable,
mixed models demonstrate robustness to missing data under the Missing at
Random (MAR) assumption and with a correct variance-covariance
specification.\cite{molenberghs2007}.

@@

Clean up your dataset before the imputation process. For imputation,
include the variables necessary for your analysis, depending on your
analysis this may be the exposure or intervention, outcome, covariates,
confounders, stratification factors, and the cluster variable. Think
about adding other useful variables, even if they're not in the main
model, such as instrumental or auxiliary variables which might improve
the accuracy of the imputation model, especially if they're associated
to the probability of missingness for some incomplete variables.

Figure out if you can directly impute incomplete variables by using
deductive imputation. This involves inferring missing values based on
logical connections between variables. It's especially useful for
variables that depend on each other, like calculating BMI from weight
and height.

Deductive imputation is also useful for getting values for level-1
variables from level-2 ones, like in Individual Participant Data (IPD).
For example, deductive imputation can be employed to infer missing test
values for patients associated with a center unit that lacks a test
procedure or machine, utilizing accessible metadata or protocol details.

@@

Start by choosing the imputation model for each incomplete variable in
your dataset. The mice package suggests methods based on variable types
for non-clustered variables, and for the clustered ones you can use the
micemd package's find.defaultMethod() function. This function selects
from different 2l imputation methods based on cluster size and the
proportion of missing data in each cluster.

@@

The imputation model can incorporate additional terms compared to the
main model without causing compatibility issues. For instance, it's
recommended to include the outcome variable in the imputation model for
prediction variables \cite{moons2006a}. In cases where the outcome is
time-to-event, the Nelson-Aalen estimate of the time to the event should
be added as a covariate in the imputation model, as well as the event
indicator {[}REF: White IR, Royston P. Imputing missing covariate values
for the Cox model. Stat Med. 2009 Jul 10;28(15):1982-98. Available from:
doi.org/10.1002/sim.3618{]}. Additionally, including auxiliary
variables, even if not part of the main model, can be linked to the
probability of missingness, improving the likelihood of meeting the
Missing at Random (MAR) assumption and enhancing estimation efficiency
\cite{hardt2012a}.

@@

\begin{tabular}{ll}
     \toprule
Type & Definition \\
  \midrule
 1  & Fixed variable, e.g., level-1  $x_{1_{ij}}$ or  level-2 $x_{2_{i.}}$ \\
 2  & Random variable, e.g., $x_{3_{ij}}$\\
 3  & *Fixed variable with cluster mean $x_{4_{ij}}$\\
 4  & *Random variable with cluster mean $x_{5_{ij}}$\\
  \midrule
-2  & Cluster variable, in this case the one defined by j index  \\
-3. & Random variable only included on selection model (Heckman model)\\
-4. & Random variable only included on main model (Heckman model)\\
      \bottomrule
\end{tabular}

\begin{itemize}
\tightlist
\item
  It has been advised to use variables of type 3 and 4, as the inclusion
  of the means of the cluster is beneficial on FCS \cite{mistler2017}.
\end{itemize}

Recipes have been proposed for imputing incomplete level-1 and level-2
variables for hierarchical models \cite{buuren2018a} (\(\S\) 7.10),
which can be convenient to follow when dealing with numerous variables
(interaction terms at different levels) and models with many random
effects that are prone to convergency problems due to overspecification
in the imputation model. These rules were designed to ensure
compatibility among the conditionally specified imputation models and
congeniality between the imputation and the main model. However, they
may not be applicable to all 2l imputation methods. For instance, the
2l.2stage imputation methods of the micemd package only allow the
inclusion of random predictor variables (2).

@@

\hypertarget{sec-impact}{%
\subsection{IMPACT data}\label{sec-impact}}

@@

In this example, we demonstrate a multilevel imputation of random
intercept and random slope model with a continuous response. We utilize
the obesity dataset included in the \texttt{micemd}@ package, a
simulated dataset that emulates an electronic survey in which
individuals are asked to provide information about their weight and
consumption habits in different countries. We use the following
variables from the dataset:

@@

We observe that the missing pattern is non-monotonic and quite similar
across the clusters. However, regarding the weight variable, we notice
that it is systematically missing in cluster 3. In order to evaluate if
we require a imputation method that accounts for clustering we assess
the Intraclass Correlation

@@

It suggests using `2l.2stage.bin' for the FamOb variable and
`2l.2stage.norm' for the age variable. However, after inspecting the age
density plot, we consider modifying its method to `2l.2stage.pmm'. For
the BMI variable, we employ deterministic imputation.

@@

For these imputation models, it is necessary to specify the prediction
matrix, with the cluster variable labelled as -2 and the predictor
variable measured within clusters labelled as 2, encompassing all
variables. We need to supress the variable Time as this variable is not
specified in the main model. We also modify the relationship between
BMI, weight and height in the prediction matrix to avoid circular
predictions. Then we proceed to run the imputation model.

@@

After confirming convergence, we proceed to save the results for future
use.

@@

We consider the possibility that patients may not have been selected
randomly, which would then have led to a distribution for weight that
does not reflect the weight in the population. It's likely that an
omitted variable, like self-esteem, could influence this selection. For
instance, individuals with lower self-esteem might have higher weight
values, impacting their willingness to provide honest information due to
embarrassment.

To address this situation where data are Missing Not At Random (MNAR),
one approach is to apply the Heckman selection model. This method has
recently been extended to allow for variations in intercepts and
exposure effects (random intercept and slope)
\cite{galimard2018, munoz2023}. This method involves specifying two
equations: one for the outcome, describing the incomplete variable in
terms of partially observed predictors (in this case, all variables from
the main model), and the other for the selection model, explaining the
probability of being observed based (R) on certain variables.

To apply the \textbf{2l.2stage.heckman} method, the weight variable
should be specified as `2l.2stage.heckman' found in the micemd package.
Additionally, the prediction matrix needs modification, as this model
includes one submodel for the individuals' measured outcome, and one for
the selection of individuals. For the outcome equation we consider the
same imputation model that we used for the MAR case (main model).
\[Weight_{ij}= \beta^O_{o} + \beta^O_{1}Age_{ij} + \beta^O_{2}FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}\]
Regarding the selection equation, we include the same predictors as
those in the main model, as well as a time variable. Here the time
variable serves as a restriction exclusion variable specifically
explaining the probability of being observed but not affecting the
incomplete value (Weight). In this context, we assume that the time a
user spends completing the survey serves as a proxy for the barriers
they may encounter in survey completion, such as familiarity with the
survey content or internet speed. These factors may lead the user to
skip specific questions or even the entire survey. Also, we assume the
time does not have any influence on the subject's weight.
\[R_{ij}= \beta^S_{o} + \beta^S_{1}Age_{ij} + \beta^S_{2}FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}\]

@@

We note that there is minimal disparity in the age effect, FamObs, or
Gender across the various imputation models under consideration. An
analysis of the intercept reveals that, under the MNAR assumption, a
higher average BMI is anticipated compared to the MAR assumption.
Nonetheless, with respect to precision of the estimates, we notice that
in general assuming MNAR leads to wider confidence intervals. In this
case it does not have any influence on the final result but there could
be cases where variation in the assumed missing mechanism could lead
also to differences on the significance of a statistical test and
therefore lead to different conclusions.

@@

The ``Mice'' and ``Mice''-based R packages offer a range of imputation
methods tailored for hierarchical data, easily adaptable to the
dataset's structure. Before proceeding with the analysis of the imputed
dataset, it is essential to assess the convergence of the imputation
method. This evaluation can reveal issues such as circular problems, the
need for additional iterations, or challenges associated with the chosen
imputation method.

@@

\hypertarget{disclaimer}{%
\section{Disclaimer}\label{disclaimer}}
